R Notebook

### 2018-06-13 SNP_sWGS_comparison.Rmd
adding script to process the SNP array data from Reid et.al.

```{r, include=FALSE}
library(copynumber)
library(tibble)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(glmnet)

suppressPackageStartupMessages( source('lib/data_func.R') )
```

@@

grid.arrange(
  ggplot(relCNV, aes(x=1:nrow(relCNV), y=D702_D506_10725_10729)) + geom_point(color='darkblue') + labs(title='Raw', x=''),
  ggplot(melt(segCNV['D702_D506_10725_10729',]), aes(x=1:ncol(segCNV), y=value)) + geom_point(color='darkblue') + labs(title='Tiled & Logged', y='D702_D506_10725_10729 (IMC)', x='') )

grid.arrange(
  ggplot(relCNV, aes(x=1:nrow(relCNV), y=D706_D507_9242)) + geom_point(color='darkblue') + labs(title='Raw', x=''),
  ggplot(melt(segCNV['D706_D507_9242',]), aes(x=1:ncol(segCNV), y=value)) + geom_point(color='darkblue') + labs(title='Tiled & Logged', y='D706_D507_9242 (NDBE)', x='') )

grid.arrange(  
  ggplot(relCNV, aes(D702_D506_10725_10729)) + geom_histogram(color='lightblue', bins=25) + labs(title='Raw'),
  ggplot(melt(unlogged['D702_D506_10725_10729',]), aes(value)) + geom_histogram(color='lightblue', bins=25) + labs(title='Segmented'),
  ggplot(melt(segCNV['D702_D506_10725_10729',]), aes(value)) + geom_histogram(color='lightblue', bins=25) + labs(title='Logged & Segmented'), top='IMC'
)

grid.arrange(  
  ggplot(relCNV, aes(D706_D507_9242)) + geom_histogram(color='lightblue', bins=25) + labs(title='Raw'),
  ggplot(melt(unlogged['D706_D507_9242',]), aes(value)) + geom_histogram(color='lightblue', bins=25) + labs(title='Segmented'),
  ggplot(melt(segCNV['D706_D507_9242',]), aes(value)) + geom_histogram(color='lightblue', bins=25) + labs(title='Logged & Segmented'), top='NDBE'
)

# All samples
load('~/Data/Ellie/Analysis/5e6_arms_all_logR/model_data.Rdata', verbose=T)
load('~/Data/Ellie/Analysis/5e6_arms_all_logR/loo.Rdata', verbose=T)
subset(pg.samp$AH0254, Samplename %in% c('D706_D507_9242','D702_D506_10725_10729'))

@@

GSE99429 LogR data
```{r, warning=F, message=F}
filelist = list.files('~/Data/Ellie/arrays/SNP', pattern='txt', full.names=T)
samplelist = lapply(filelist, function(f) colnames(as_tibble(data.table::fread(f, nrows=2))))

whole_epi = lapply(samplelist, function(s) grep('*whole.*Log', s, value=T)  )
names(whole_epi) = basename(filelist)

normals = lapply(samplelist, function(s) grep('*normal.*Log', s, value=T)  )
names(normals) = basename(filelist)

@@

dt = as_tibble(data.table::fread(filelist[6]))
epi = dt[,c(grep('SNP|Chr|Pos', colnames(dt), value=T),whole_epi[[6]])]
nm = dt[,c(grep('SNP|Chr|Pos', colnames(dt), value=T),normals[[6]])]

epi = fix.chr(epi)
nm = fix.chr(nm)
rm(dt)

@@

```{r, warning=F, message=F}

@@

Normal sample?
```{r, warning=F, message=F}

@@

Norm vs whole ep
The extreme tail on the normal sample is due to the 'cx' coefficient and needs to be adjusted.  For now I will set it to the mean of the cx value in the sWGS dataset.
```{r}

@@

Compare to all samples.
```{r warning=F, message=F}

@@

Quantile normalize?
```{r, warning=F, message=F}
by = 3
mx = mean(arrayDf, na.rm=T)+sd(arrayDf, na.rm=T)*by
mn = mean(arrayDf, na.rm=T)-sd(arrayDf, na.rm=T)*by

@@

## Try processing the IDAT files directly.  Perhaps if they aren't normalized together the LogR becomes more comparable.
```{r}

library(illuminaio)

file = '~/Data/Ellie/arrays/SNP/GSM2643883_8832987072_R03C01_Grn.idat'
idat <- readIDAT(file)
names(idat)

>>>>

```{r, include=FALSE}
library(copynumber)
library(tibble)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(glmnet)
library(reshape2)

winsor2<-function (x, multiple=3) {
   if(length(multiple) != 1 || multiple <= 0) {
      stop("bad value for 'multiple'")
   }
   med <- median(x)
   y <- x - med
   sc <- mad(y, center=0) * multiple
   y[ y > sc ] <- sc
   y[ y < -sc ] <- -sc
   y + med
}

suppressPackageStartupMessages( source('lib/data_func.R') )
load('~/Data/Ellie/Analysis/5e6_arms_all_logR/model_data.Rdata', verbose = T)
load('~/Data/Ellie/Analysis/5e6_arms_all_logR/all.pt.alpha.Rdata', verbose = T)
fitV = models$`0.9`
l = performance.at.1se$`0.9`$lambda
rm(plots,performance.at.1se,dysplasia.df,models,cvs,labels)

@@

chr.info = get.chr.lengths()[1:22,]
chr.info$chrom = sub('chr', '', chr.info$chrom)
chr.info$chrom = factor(chr.info$chrom, ordered = T, levels=c(1:22))
relCNV$chrom = factor(relCNV$chrom, ordered = T, levels = c(1:22))
(segCNV['D708_D504_9242',])

windows = apply(chr.info, 1, function(x) {
  m = data.frame(matrix(seq(from=5e6,to=x[['chr.length']],by=5e6),ncol=1))
  cbind(x[['chrom']], m)
})
names(windows) = c(1:22)
windows = do.call(rbind,windows)
colnames(windows) = c('chrom','window')

# tiles = tile.segmented.data(relCNV, chr.info=chr.info, verbose=T)
# tiles.arms = tile.segmented.data(relCNV, size='arms',chr.info=chr.info, verbose=T)

# ggplot(subset(chr.info, chrom == 2), aes(x=1:chr.length)) + #facet_grid(~chrom,space = 'free_x', scales = 'free_x') + 
#   labs(x='Chromosome 2', y='Segmented values') +
#   geom_segment(data=subset(relCNV[,c('chrom','start.pos','end.pos','D708_D504_9242')], chrom==2), aes(x=start.pos, xend=end.pos, y=D708_D504_9242, yend=D708_D504_9242), size=3) +
#   geom_vline(data=subset(windows, chrom==2), aes(xintercept=window), linetype='dashed', color='grey39') +
#   geom_segment(data=subset(tiles, chr == 2), aes(x=start, xend=end, y=D708_D504_9242, yend=D708_D504_9242), color='red', size=2) +
#   geom_segment(data=subset(tiles.arms, chr == 2), aes(x=start, xend=end, y=D708_D504_9242, yend=D708_D504_9242), color='blue', size=2)

# IMC
grid.arrange(  
  ggplot(relCNV, aes(D708_D504_9242)) + geom_histogram(color='lightblue', bins=25) + labs(title='Raw'),
  ggplot(melt(unlogged['D708_D504_9242',]), aes(value)) + geom_histogram(color='lightblue', bins=25) + labs(title='Segmented'),
  ggplot(melt(segCNV['D708_D504_9242',]), aes(value)) + geom_histogram(color='lightblue', bins=25) + labs(title='Logged & Segmented'), top='IMC'
)

qqnorm(segCNV['D708_D504_9242',])
qqnorm(unlogged['D708_D504_9242',])
qqnorm(relCNV[,'D708_D504_9242']) # raw

# Early NDBE
grid.arrange(  
  ggplot(relCNV, aes(D706_D507_9242)) + geom_histogram(color='lightblue', bins=25) + labs(title='Raw'),
  ggplot(melt(unlogged['D706_D507_9242',]), aes(value)) + geom_histogram(color='lightblue', bins=25) + labs(title='Segmented'),
  ggplot(melt(segCNV['D706_D507_9242',]), aes(value)) + geom_histogram(color='lightblue', bins=25) + labs(title='Logged & Segmented'), top='NDBE'
)

qqnorm(segCNV['D706_D507_9242',])
qqline(segCNV['D706_D507_9242',])

qqnorm(unlogged['D706_D507_9242',])
qqnorm(relCNV[,'D706_D507_9242'])

subset(pg.samp$AH0254, Samplename %in% c('D706_D507_9242','D702_D506_10725_10729'))

@@

GSE99429 LogR data
```{r, warning=F, message=F, eval=F}
filelist = list.files('~/Data/Ellie/arrays/SNP', pattern='txt', full.names=T)

whole_epiP = as_tibble(data.table::fread(grep('391_62K.*whole_epi', filelist, value=T)[1], sep='\t', header=T))
#normalP = as_tibble(data.table::fread(grep('391.*normal.*200ng', filelist, value=T), sep='\t', header=T))

#whole_epiNP = as_tibble(data.table::fread(grep('451_115T.*whole_epi', filelist, value=T)[1], sep='\t', header=T))
#normalNP = as_tibble(data.table::fread(grep('451.*normal.*200ng', filelist, value=T), sep='\t', header=T))

@@

whole_epiP = fix.chr(whole_epiP)
normalP = fix.chr(normalP)
qqnorm(whole_epiP$`391_62K_SS4_whole_epi__22325.Log.R.Ratio`)
#qqline(whole_epiP$`391_62K_SS4_whole_epi__22325.Log.R.Ratio`)

head(whole_epiP)

chrs = get.chr.lengths()[1:22,]
chrs$chrom = factor(sub('chr','',chrs$chrom), ordered = T, levels=c(1:22))
colnames(chrs)[1] = 'Chr'

whole_epiP$Chr = factor(whole_epiP$Chr, levels=c(1:22), ordered = T)

p = ggplot(chrs, aes(x=1:chr.length)) + facet_grid(~Chr, space='free_x', scales='free_x') +
  geom_point(data=whole_epiP, aes(Position, `391_62K_SS4_whole_epi__22325.Log.R.Ratio`))

#qqnorm(whole_epiP$`391_62K_SS4_whole_epi__22325.Log.R.Ratio`)
#epiW = winsorize(as.data.frame(whole_epiP[,c(1:3)]))
#head(epiW)

whole_epiP$Chr = as.integer(as.character(whole_epiP$Chr))
#nas = which(is.na(whole_epiP$`391_62K_SS4_whole_epi__22325.Log.R.Ratio`))

seg = pcf(as.data.frame(whole_epiP[,c(1:3)]), gamma=37, verbose=T)
head(seg)

x = which(seg$chrom == 17)[1]

baf = whole_epiP[,c(1,2,4)]
summary(subset(baf, Position >= seg$start.pos[x] & Position <= seg$end.pos[x] & Chr == seg$chrom[x])[,3,drop=T], na.rm=T)

summary(seg$end.pos-seg$start.pos)
summary(seg$mean)
colnames(seg)[2] = 'Chr'
seg$Chr = factor(seg$Chr, levels=levels(chrs$Chr), ordered = T)
head(seg)
p + geom_segment(data=seg, aes(x=start.pos,xend=end.pos, y=mean, yend=mean), color='red', size=2)

#hist(seg$mean)

seg$meanW = NA
seg$meanW[!is.na(seg$mean)] = winsor2(seg$mean[!is.na(seg$mean)]) # winsorize w/replacement

tiled = tile.segmented.data(seg[,c(2:6,8)], chr.info = get.chr.lengths(), verbose = T)
head(tiled)
hist(tiled$meanW)
# colnames(tiled)[1] = 'Chr'
# tiled$Chr = factor(tiled$Chr, ordered = T)
# p + geom_segment(data=tiled, aes(x=start,xend=end, y=mean, yend=mean), color='red', size=2)

colnames(tiled)[4] = 'mean'
p1 = ggplot(tiled) + geom_histogram(aes(mean), col='lightblue', bins=20) + 
  geom_vline(xintercept=c(mean(tiled$mean,na.rm=T)+sd(tiled$mean,na.rm=T), mean(tiled$mean,na.rm=T)-sd(tiled$mean,na.rm=T)), col='red' )  
tiled$scaled = scale(exp(tiled$mean))
p2 = ggplot(tiled) + geom_histogram(aes(scaled), col='lightblue', bins=20) +
    geom_vline(xintercept=c(mean(tiled$scaled,na.rm=T)+sd(tiled$scaled,na.rm=T), mean(tiled$scaled,na.rm=T)-sd(tiled$scaled,na.rm=T)), col='red' )  
grid.arrange(p1, p2)

qqnorm(tiled$scaled)
hist(tiled$scaled)

#x = tiled

tiled = segment.matrix(tiled)
# for (i in 1:ncol(tiled)) 
#   tiled[,i] = unit.var(tiled[,i], z.mean[i], z.sd[i]) 
tiled[is.na(tiled)] = mean(tiled, na.rm=T) #0

tiled.arms = tile.segmented.data(seg[,c(2:6,8)], size='arms', chr.info=get.chr.lengths(), verbose=T)
colnames(tiled.arms)[4] = 'mean'
tiled.arms$scaled = scale(exp(tiled.arms$mean))
#tiled.arms$mean = scale(exp(tiled$mean))
tiled.arms = segment.matrix(tiled.arms)
tiled.arms[is.na(tiled.arms)] = mean(tiled.arms, na.rm=T) #0

predict(fitV, newx=arrayDf, s=l, type='response')
predict(fitV, newx=arrayDf, s=l, type='link')

summary(arrayDf[1,])

hist(allDf[210,], breaks=20)

# logr = whole_epiP$`391_62K_SS4_whole_epi__22325.Log.R.Ratio`
# hist(logr)
# by = 3
# mx = mean(logr,na.rm=T)+sd(logr,na.rm=T)*by
# mn = mean(logr,na.rm=T)-sd(logr,na.rm=T)*by
# abline(v=c(mx, mn), col='red')
# 
# newsd = sd(logr[logr <= mx & logr >= mn], na.rm=T)
# newmean = mean(logr[logr <= mx & logr >= mn], na.rm=T)

# xx = qnorm(pnorm(logr, newmean, newsd, lower.tail=F))
# xx[xx==Inf] = NA
# xx[xx==-Inf] = NA
# hist(xx)
# qqnorm(xx)

epiW = winsorize(as.data.frame(whole_epiP[,c(1:3)]))
qqnorm(epiW$`391_62K_SS4_whole_epi__22325.Log.R.Ratio`)

logrw = epiW$`391_62K_SS4_whole_epi__22325.Log.R.Ratio`
hist(logrw)
by = 3
mx = mean(logrw,na.rm=T)+sd(logrw,na.rm=T)*by
mn = mean(logrw,na.rm=T)-sd(logrw,na.rm=T)*by
abline(v=c(mx, mn), col='red')

newsd = sd(logrw[logrw <= mx & logrw >= mn], na.rm=T)
newmean = mean(logr[logrw <= mx & logrw >= mn], na.rm=T)

xy = qnorm(pnorm(logrw, newmean, newsd, lower.tail=F))
xy[xy==Inf] = NA
xy[xy==-Inf] = NA
hist(xy)
qqnorm(xy)

length(xy)

rows = which(logrw <= mx & logrw >= mn)
head(epiW[rows,])
head(xy[rows])

epiW = cbind(epiW, xy)

seg = pcf(epiW[,c(1:3)], gamma=40, fast=T, verbose=F)
qqnorm(seg$mean)
hist(seg$mean)
by = 3
mx = mean(seg$mean,na.rm=T)+sd(seg$mean,na.rm=T)*by
mn = mean(seg$mean,na.rm=T)-sd(seg$mean,na.rm=T)*by
abline(v=c(mx, mn), col='red')

newsd = sd(seg$mean[ seg$mean >= mn], na.rm=T)   
newmean = mean(seg$mean[ seg$mean >= mn], na.rm=T)

xy = qnorm(pnorm(seg$mean, newmean, newsd, lower.tail=F))
xy[xy==Inf] = NA
xy[xy==-Inf] = NA
hist(xy)
qqnorm(xy)

summary(seg$end.pos-seg$start.pos)

chrs = get.chr.lengths()
chrs$chrom = sub('chr','',chrs$chrom)

sc = base::merge(seg, chrs, by='chrom')
head(sc)

unseg = base::merge(epiW, chrs, by='chrom')
head(unseg)
unseg = cbind(unseg, xy)

unseg = unseg %>% rowwise() %>% dplyr::mutate(g.pos = ifelse(chrom == '1', pos, genome.length-pos))

p = ggplot(unseg) + xlim(1, max(chrs$genome.length)) + 
  geom_point(aes(x=g.pos, y=xy), color='red', alpha=0.5) +
  geom_point(aes(x=g.pos, y=`391_62K_SS4_whole_epi__22325.Log.R.Ratio`)) 
ggsave(p, width=10, height = 7, units = 'in', file='snps.png')

sc = sc %>% rowwise() %>% dplyr::mutate(g.start = ifelse(chrom == '1', start.pos, genome.length-start.pos), 
                                        g.end = ifelse(chrom == '1', end.pos, genome.length-end.pos))

p + geom_segment(data=sc, aes(x=g.start, xend=g.end, y=mean, yend=mean), color='red', size=2)

ggplot() + lims(x=c(1,max(chrs))) + 
  geom_point(data=epiW)

ggplot(sc) + facet_grid(~chrom, space = 'free_x', scales = 'free_x') + geom_segment(aes(x=start.pos, xend=end.pos, y=mean, yend=mean))

#x = imputeMissing(epiW, 'constant')
#asseg = aspcf(x, epibaf, gamma=40, kmin=64)

#segdata = copynumber::pcf(as.data.frame(data), gamma=28, fast=T, verbose=F)

@@

```{r, warning=F, message=F, eval=F}

@@

Normal sample?
```{r, warning=F, message=F, eval=F}

@@

Norm vs whole ep
The extreme tail on the normal sample is due to the 'cx' coefficient and needs to be adjusted.  For now I will set it to the mean of the cx value in the sWGS dataset.
```{r, eval=F}

@@

Compare to all samples.
```{r warning=F, message=F, eval=F}

@@

Quantile normalize?
```{r, warning=F, message=F, eval=F}
by = 3
mx = mean(arrayDf, na.rm=T)+sd(arrayDf, na.rm=T)*by
mn = mean(arrayDf, na.rm=T)-sd(arrayDf, na.rm=T)*by

@@

Try processing the IDAT files directly.  Perhaps if they aren't normalized together the LogR becomes more comparable?

## Absolutely CN

```{r, warning=F, message=F, echo=F}
ploidy = read.table('~/Data/ICGC/BE_OAC_Cohort/tumour/data/ascat/LP2000104-DNA_A01_vs_LP2000101-DNA_A01/LP2000104-DNA_A01_vs_LP2000101-DNA_A01__pv.1.5__rg.grch37_g1k__al.bwa_mem__.samplestatistics.csv', row.names=1)['Ploidy',1]
cn = data.table::fread('~/Data/ICGC/BE_OAC_Cohort/tumour/data/ascat/LP2000104-DNA_A01_vs_LP2000101-DNA_A01/LP2000104-DNA_A01_vs_LP2000101-DNA_A01__pv.1.5__rg.grch37_g1k__al.bwa_mem__.copynumber.caveman.csv')
cn = subset(cn, Chromosome %in% c(1:22))

cn = read.table('~/Data/ICGC/BE_OAC_Cohort/BA/data/ascat/totalCNV.txt', sep='\t', header=T)
cn = subset(cn, Illumina.Barcode == 'LP2000110-DNA_A01_vs_LP2000101-DNA_A01')
ploidy = cn$ploidy[1]
nrow(cn)

x = cnN$Total_CN-ploidy
qqnorm(x)
qqnorm(winsor2(x, 3))

cn[which(cn$Chromosome == 17)[1],]
x[which(cn$Chromosome == 17)[1]]
winsor2(x, 3)[which(cn$Chromosome == 17)[1]]

cn$TotalWinsAdj = winsor2(x)

cols = grep('Chr|Start|End|TotalWinsAdj', colnames(cn), ignore.case=T)

tiled = tile.segmented.data(as.data.frame(cn)[cols], chr.info=get.chr.lengths()[1:22,])
tiled.arms = tile.segmented.data(as.data.frame(cn)[cols], size='arms', chr.info=get.chr.lengths()[1:22,])

tt = segment.matrix(tiled)
ta = segment.matrix(tiled.arms)

tt[is.na(tt)] = mean(tt, na.rm=T) #0
ta[is.na(ta)] = mean(ta, na.rm=T) #0

cx = score.cx(tt, 1)

dm = subtract.arms(tt, ta)

dm = cbind(dm, 'cx'=unit.var(cx, mn.cx, sd.cx ))

#unit.var(cx)
predict(fitV, newx=dm, s=l, type='response')
predict(fitV, newx=dm, s=l, type='link')

## in silico normal sample
cnN = cn
cnN$Total_CN = sample(c(1,2,3), size=nrow(cnN), replace=T, prob=c(.025,.95,.025 ))
hist(cnN$Total_CN)

cnN$TotalWinsAdj = winsor2(cnN$Total_CN-round(ploidy))
qqnorm(cnN$TotalWinsAdj)

hist(cnN$TotalWinsAdj)

cols = grep('Chr|Start|End|TotalWinsAdj', colnames(cn), ignore.case=T)

tiled = tile.segmented.data(as.data.frame(cn)[cols], chr.info=get.chr.lengths()[1:22,])
tiled.arms = tile.segmented.data(as.data.frame(cn)[cols], size='arms', chr.info=get.chr.lengths()[1:22,])

tt = segment.matrix(tiled)
ta = segment.matrix(tiled.arms)

tt[is.na(tt)] = mean(tt, na.rm=T) #0
ta[is.na(ta)] = mean(ta, na.rm=T) #0

cx = score.cx(tt, 1)

dm = subtract.arms(tt, ta)

dm = cbind(dm, 'cx'=unit.var(cx, mn.cx, sd.cx ))

#unit.var(cx)
predict(fitV, newx=dm, s=l, type='response')
predict(fitV, newx=dm, s=l, type='link')

hist(allDf[1,])
